---
title: "Creating a basic data structure (BDS) Exposure ADaM"
author: "Lun-Hsien Chang"
date: "01-Dec-2024"
toc: true
toc-depth: 2  # Levels of headings to include in TOC (e.g., 2 includes H1 and H2)
toc-title: "Programming workflow"  # Custom title for the TOC
format:
  html:
    css: styles.css  # Use custom CSS for soft-wrap in HTML
  pdf: 
    latex_engine: xelatex
    header-includes:
      - \usepackage{listings}
      - \lstset{breaklines=true, breakatwhitespace=true, basicstyle=\ttfamily\small, columns=fullflexible}
    fontsize: 11pt
    geometry: margin=1in, left=1in, right=1in, top=1in, bottom=1in
    mainfont: "Times New Roman"
    fontenc: T1
    citation_package: natbib
    urlcolor: blue       # Make URLs blue
    linkcolor: blue      # Make links in the PDF blue
    hyperref: true
  markdown:
    variant: gfm         # Use GitHub Flavored Markdown
    embed-resources: true # Embed images and other resources
output-dir: "outputs"
editor: visual
source: "https://cran.r-project.org/web/packages/admiral/vignettes/bds_exposure.html"
---

Get required R packages

```{r}
#| echo: false
library(admiral)
library(dplyr, warn.conflicts = FALSE)
library(pharmaversesdtm)
library(lubridate)
library(stringr)
library(tibble)
```

# Read CDISC pilot SDTM and ADaM datasets

```{r}
#| echo: false
data("admiral_adsl")
data("ex")

adsl <- admiral_adsl # dim(adsl) 306 50
ex <- convert_blanks_to_na(ex) # dim(ex) 591 17
```

```{r}
#| wrap: true
adsl_vars <- exprs(TRTSDT, TRTSDTM, TRTEDT, TRTEDTM)

# left join EX and adsl TRTSDT, TRTSDTM, TRTEDT, TRTEDTM on ex.STUDYID=adslSTUDYID and ex.USUBJID=adsl.USUBJID
adex <- derive_vars_merged(
   dataset=ex
  ,dataset_add = adsl
  ,new_vars = adsl_vars
  ,by_vars = exprs(STUDYID, USUBJID)
  ) # dim(adex) 591 21
```

The CDISC pilot `EX` domain data does not contain a dose adjustment flag or the planned dose information. For demonstration purposes, this will be added to the data.

## EXADJ

-   Exposure Adjustment ?

## EXDOSE

-   exposure dose

-   from SDTM.EX.EXDOSE

## EXPLDOS

-   Planned Dose

```{r}
adex <- adex %>%
  mutate(
    EXADJ = case_when(
      USUBJID == "01-701-1028" & VISIT %in% c("WEEK 2") ~ "ADVERSE EVENT",
      USUBJID == "01-701-1148" & VISIT %in% c("WEEK 2", "WEEK 24") ~ "MEDICATION ERROR",
      TRUE ~ NA_character_
    ),
    EXDOSE = case_when(
      USUBJID == "01-701-1028" & VISIT %in% c("WEEK 2") ~ 0,
      USUBJID == "01-701-1148" & VISIT %in% c("WEEK 2", "WEEK 24") ~ 0,
      TRUE ~ EXDOSE
    )
  ) %>%
  mutate(EXPLDOS = if_else(EXTRT == "PLACEBO", 0, 54))

adex %>% select(EXTRT, EXPLDOS) %>% distinct()
```

# Derive numeric datetime, analysis day variables

## ASTDT

-   Analysis Start Date

-   Set to a numeric form of EX.EXSTDTC when EX.EXSTDTC consists of a full date.

## AENDT

-   Analysis End Date

-   Set to a numeric form of EX.EXENDTC when EX.EXENDTC consists of a full date.

```{r}
# Convert character datetime to numeric datetime
adex <- derive_vars_dt(adex, new_vars_prefix = "AST", dtc = EXSTDTC)
adex <- derive_vars_dt(adex, new_vars_prefix = "AEN", dtc = EXENDTC) # dim(adex) 591 25
adex %>% select(USUBJID,VISIT,EXSTDTC,EXENDTC,ASTDT,AENDT) %>% head()
```

## ASTDTM

## AENDTM

The next examples demonstrates the datetime imputation features available in the `derive_vars_dtm()` function, where the time is imputed as “00:00:00”:

```{r}
adex <- derive_vars_dtm(
  adex
  ,dtc = EXSTDTC
  # Impute dtc date to the first day of the month
  ,highest_imputation = "M"
  ,date_imputation = "first"
  ,new_vars_prefix = "AST"
)

adex <- derive_vars_dtm(
  adex,
  dtc = EXENDTC,
  # Impute dtc date to the last day of the month
  highest_imputation = "M",
  date_imputation = "last",
  new_vars_prefix = "AEN"
)

adex %>% select(EXSTDTC,EXENDTC,ASTDTM,AENDTM) %>% head()
```

## ASTDY

-   Analysis Start Day

-   \`ASTDT-TRTSDT+1\`

## AENDY

-   Analysis End Day

-   \`AENDT-TRTSDT+1\`

```{r}
adex <- derive_vars_dy(
  dataset=adex
  ,reference_date = TRTSDT
  ,source_vars = exprs(ASTDT, AENDT)
  ) # dim(adex) 591 33
adex %>% select(TRTSDT, ASTDT, ASTDY, AENDT, AENDY) %>% head()
```

## EXDURD

-   Duration of treatment or exposure

-   \`EXDURD=AENDT - ASTDT +1\`

```{r}
adex <- adex %>%
  derive_vars_duration(
    new_var = EXDURD
    ,start_date = ASTDT
    ,end_date = AENDT 
    # duration unit can be "years", "months", "weeks", "days", "hours", "minutes", "seconds"
    ,out_unit = "DAYS")

adex %>% select(ASTDT, AENDT, EXDURD) %>% head()
```

## DOSEO

-   Dose Overall (?). Refers to the actual dose of the study drug administered, standardized (e.g., per unit of body weight, such as mg/kg).

-   \``EXDOSE * EXDURD`\`

## PDOSEO

-   Planned Dose Overall (?) Represents the planned dose per protocol, similarly normalized
-   \`EXPLDOS \* EXDURD\`

```{r}
adex <- adex %>%
  mutate(
    DOSEO = EXDOSE * EXDURD
    ,PDOSEO = EXPLDOS * EXDURD)

adex %>% select(USUBJID, EXDOSE, EXPLDOS, EXDURD, DOSEO, PDOSEO) %>% head()
```

## Create 1:1 mapping records

The first set of exposure records to create will be records mapped 1:1 to an existing collected exposure record in SDTM. For these records, the `AVAL` or `AVALC` would be calculated using columns that exist on the data and no summarizing of records would be necessary.

These records may be used for input into summary records or be used individually for summarization in outputs. Some examples may be exposure duration, dose administered, dose adjusted, etc. based on one exposure record in SDTM.

These records can be derived using simple `dplyr::mutate` assignments and then combined

## PARAMCD

-   Parameter Code

## AVALC

-   Analysis Value Character ?

```{r}
adex_durd <- adex %>%
  mutate(
    PARAMCD = "DURD",
    AVAL = EXDURD)

adex_dose <- adex %>%
  mutate(
    PARAMCD = "DOSE",
    AVAL = DOSEO)

adex_pldos <- adex %>%
  mutate(
    PARAMCD = "PLDOSE",
    AVAL = PDOSEO)

adex_adj <- adex %>%
  mutate(
    PARAMCD = "ADJ",
    AVALC = if_else(!is.na(EXADJ), "Y", NA_character_)
  )

adex_adjae <- adex %>%
  mutate(
    PARAMCD = "ADJAE",
    AVALC = if_else(EXADJ == "ADVERSE EVENT", "Y", NA_character_)
  )

adex <- bind_rows(
  adex_durd,
  adex_dose,
  adex_pldos,
  adex_adj,
  adex_adjae) %>%
  mutate(PARCAT1 = "INDIVIDUAL") # dim(adex) 2955 40

adex %>% select(USUBJID,VISIT,ASTDT,AENDT,PARAMCD,AVAL,AVALC) %>% head()
```

## Create Summary Records

Exposure is commonly analyzed by a timing interval (e.g. `APHASE`, `APERIOD`, `AVISIT`, etc.). For these types of calculations, the `derive_param_exposure()` function may be used. In addition to creating a summarized `AVAL`, the function will also compute minimum and maximum dates for the record. For example, to calculate the total dose by subject and treatment

```{r}
adex <- derive_param_exposure(
  dataset = adex
  ,dataset_add = adex
  ,by_vars = exprs(STUDYID, USUBJID, !!!adsl_vars)
  ,input_code = "DOSE"
  ,set_values_to = exprs(
    PARAMCD = "TDOSE",
    PARCAT1 = "OVERALL",
    AVAL = sum(AVAL, na.rm = TRUE)
  )
) # dim(adex) 3209 40
```

# References

[Creating a BDS Exposure ADaM](https://cran.r-project.org/web/packages/admiral/vignettes/bds_exposure.html)

[ADaM Subject-level Analysis - ADSL Dataset](https://www.cdisc.org/kb/examples/adam-subject-level-analysis-adsl-dataset-80283806)
