---
title: "Creating a basic data structure (BDS) Time-to-Event ADaM"
author: "Lun-Hsien Chang"
date: "04-Jan-2025"
toc: true
toc-depth: 2  # Levels of headings to include in TOC (e.g., 2 includes H1 and H2)
toc-title: "Programming workflow"  # Custom title for the TOC
format:
  html:
    css: styles.css  # Use custom CSS for soft-wrap in HTML
  pdf: 
    latex_engine: xelatex
    header-includes:
      - \usepackage{listings}
      - \lstset{breaklines=true, breakatwhitespace=true, basicstyle=\ttfamily\small, columns=fullflexible}
    fontsize: 11pt
    geometry: margin=1in, left=1in, right=1in, top=1in, bottom=1in
    mainfont: "Times New Roman"
    fontenc: T1
    citation_package: natbib
    urlcolor: blue       # Make URLs blue
    linkcolor: blue      # Make links in the PDF blue
    hyperref: true
  markdown:
    variant: gfm         # Use GitHub Flavored Markdown
    embed-resources: true # Embed images and other resources
output-dir: "outputs"
editor: visual
source: "https://cran.r-project.org/web/packages/admiral/vignettes/bds_tte.html"
---

# Introduction

This article describes creating a BDS time-to-event ADaM.

The main part in programming a time-to-event dataset is the definition of the events and censoring times. {admiral} supports single events like death or composite events like disease progression or death. More than one source dataset can be used for the definition of the event and censoring times.

Note: All examples assume CDISC SDTM and/or ADaM format as input unless otherwise specified.

# Required Packages

```{r}
#| echo: false
library(admiral)
library(dplyr, warn.conflicts = FALSE)
library(pharmaversesdtm)
library(lubridate)
```

# Read in Data

To start, all datasets needed for the creation of the time-to-event dataset should be read into the environment. This will be a company specific process.

For example purpose, the ADSL dataset—which is included in `{admiral}`—and the SDTM datasets from `{pharmaversesdtm}` are used.

```{r}
data("ae")
data("admiral_adsl")

ae <- convert_blanks_to_na(ae) # dim(ae) [1] 1191   35
adsl <- admiral_adsl # dim(adsl) [1] 306  50
```

The following code creates a minimally viable ADAE dataset to be used throughout the following examples.

```{r}
adae <- ae %>%
  left_join(adsl, by = c("STUDYID", "USUBJID")) %>%
  derive_vars_dt(
    new_vars_prefix = "AST",
    dtc = AESTDTC,
    highest_imputation = "M"
  ) %>%
  derive_vars_dt(
    new_vars_prefix = "AEN",
    dtc = AEENDTC,
    highest_imputation = "M",
    date_imputation = "last"
  ) %>%
  mutate(TRTEMFL = if_else(ASTDT >= TRTSDT &
    AENDT <= TRTEDT + days(30), "Y", NA_character_)) # dim(adae) [1] 1191   88
```

# Derive Parameters (`CNSR`, `ADT`, `STARTDT`)

## CNSR

-   Censor

-   Defines whether the event was censored for the subject within the parameter (period of observation truncated prior to event being observed). It is strongly recommended to use 0 as an event indicator and positive integers as censoring indicators. It is also recommended that unique positive integers be used to indicate coded descriptions of censoring reasons. CNSR is required for time-to-event parameters.

To derive the parameter dependent variables like `CNSR`, `ADT`, `STARTDT`, `EVNTDESC`, `SRCDOM`, `PARAMCD`, … the `derive_param_tte()` function can be used. It adds one parameter to the input dataset with one observation per subject. Usually it is called several times.

For each subject it is determined if an event occurred. In the affirmative the analysis date `ADT` is set to the earliest event date. If no event occurred, the analysis date is set to the latest censoring date.

The events and censorings are defined by the `event_source()` and the `censor_source()` class respectively. It defines

-   which observations (`filter` parameter) of a source dataset (`dataset_name` parameter) are potential events or censorings,

-   the value of the `CNSR` variable (`censor` parameter), and

-   which variable provides the date (`date` parameter).

The date can be provided as date (`--DT` variable) or datetime (`--DTM` variable).

CDISC strongly recommends `CNSR = 0` for events and positive integers for censorings. `{admiral}` enforces this recommendation. Therefore the `censor` parameter is available for `censor_source()` only. It is defaulted to `1`.

The `dataset_name` parameter expects a character value which is used as an identifier. The actual data which is used for the derivation of the parameter is provided via the `source_datasets` parameter of `derive_param_tte()`. It expects a named list of datasets. The names correspond to the identifiers specified for the `dataset_name` parameter. This allows to define events and censoring independent of the data.

# References

[Creating a BDS Time-to-Event ADaM](https://cran.r-project.org/web/packages/admiral/vignettes/bds_tte.html)

[ADaMIG V1.3](https://library.cdisc.org/browser/#/mdr/adam/adamig-1-3/datastructures/ADSL/variablesets/Identifier)
